---
title: 'Assignment 2: Exploring Storm database'
author: "C. Deramond"
output: html_document
---

## Synopsis

In this document we'll tackle the storm database for a quick exploratory data analysis. 


## Data Processing

In order to answer briefly, we'll delineate the approach:

- load the data an libraries.

```{r, cache=TRUE}
# loading libraries
library(ggplot2, warn.conflicts = F)
library(dplyr, warn.conflicts = F)
library(tidyr, warn.conflicts = F)
library(lubridate, warn.conflicts = F)
library(stringr, warn.conflicts = F)
library(data.table, warn.conflicts = F)

# data load
data <- read.csv("./data/repdata-data-StormData.csv", stringsAsFactors=FALSE)
```

Now, in order to clean the exponents, we'll join the tables by `PROPDMGEXP` and `CROPDMGEXP`, so to factor by the multiplier:


```{r, cache=TRUE}
# create small table with the factors for number conversion
exp <- data.frame(exp = c("B", "K", "M", "H"), 
                  multiplier = c(1E9, 1E3, 1E6, 1E2), 
                  stringsAsFactors = F)
# we make inner join in order to filter bad input and transform damage to real number
data <- inner_join(data, exp, by = c("PROPDMGEXP" = "exp")) %>%
        mutate(PROPDMG = PROPDMG*multiplier) %>%
        select(-multiplier, -1)
data <- inner_join(data, exp, by = c("CROPDMGEXP" = "exp")) %>%
        mutate(CROPDMG = CROPDMG*multiplier) %>%
        select(-multiplier, -1) #this should've been a function, but well...
```


Now, that we have clean date, we procceed to summarize in order to obtain our answers, following below rules:
- As per Health Cases, we define a new variable: `Total_HealthCases = FATALITIES + INJURIES` as Fatalities and number of injured people
- As per Economical Damages, we define: `Total_EconDmg = CROPDMG + PROPDMG` as the total damage is Crop lost plus Property damage.


```{r, cache=TRUE}
# moving on totals and date figures
dataFinal <- data %>%
        separate(BGN_DATE, c("date", "garbage"), sep = -8, remove = T) %>%
        mutate(date = mdy(date), year = year(date),
               EVTYPE = as.factor(EVTYPE),
               Total_HealthCases = FATALITIES + INJURIES, 
               Total_EconDmg = CROPDMG + PROPDMG) %>%
        select(year, EVTYPE, Total_HealthCases, Total_EconDmg) %>%
        group_by(EVTYPE) %>%
        summarise(HealthCases = sum(Total_HealthCases), 
                  EconDmg = sum(Total_EconDmg))  %>%
        arrange(desc(HealthCases), desc(EconDmg))
head(dataFinal)
```



## Results

As per the questions:

**1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?**


We take our dataset and order it descendently, so to take the top 10: 
```{r, cache=TRUE} 
dataFinal <- data.table(dataFinal)
head(dataFinal[order(-HealthCases)], n= 10)

```



**2. Across the United States, which types of events have the greatest economic consequences?**

We repeat the procedure in order to take top 10 from *Economical Damage*

```{r, cache=TRUE}
head(dataFinal[ order(-EconDmg)], n= 10)

```

